---
title: "Hertie School, GRAD - C6 -2001: Statistics II"
output: html_notebook
---

# Replication Project: Direct democracy and government size, evidence from Spain
## Mariana Gonzalez Carrillo, 199278
### December 16th, 2019

#### Summary

In the research paper "Direct democracy and government size, evidence from Spain",  Carlos Sanz examines the effect of direct democracy as a government system on the overall expenditures, revenues, and deficits in Spanish municipalities. He uses a regression discontinuity design and explores alternative mechanisms (reverse causality, confounded treatment by federal transfers, elite capture) and usual robustness checks for his estimations (Covariate balance, donut regressions and placebo tests at other population thresholds). For the analysis, he uses a dataset from 73.6% budgets of Spanish municipalities for the period 1978-2011.

As an alternative mechanism, I discuss the possibility of time-varying factors that might generate bias in the estimations and use a difference in differences design including a lagged variable for the treatment (direct democracy). I find that by including in the analysis more observations than those included in the bandwidths, the effect of direct democracy becomes positive and its magnitude is smaller in absolute terms. 

***

The following code is to replicate the tables  and figures from "Direct democracy and government size, evidence from Spain" by Carlos Sanz. Code developed by Mariana Gonzalez Carrillo. *Version 1. November 29th, 2019*

***
```{r}
#0. Load libraries

library(foreign)
library(tidyr)
library(kableExtra)
library(dplyr)
library(xtable)
library(stargazer)
library(lmtest)
library(zoo)
library(sandwich)
library(tidyverse)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(sjstats)
library(ggplot2)
library(rddtools)
library(moonBook)
library(ggiraph)
library(ggiraphExtra)
library(plyr)
library(rdd)
library(tidyverse)
library(rdrobust)
library(ggdag)
library(stargazer)
library(dummies)
library(KernSmooth)
library(locfit)
library(datasets)
library(RColorBrewer)
library(devtools)
install_github("MatthieuStigler/RDDtools", subdir = "RDDtools")
library(RDDtools)
library(plm)
if (!require("devtools"))
  install.packages("devtools")
devtools::install_github("jgabry/QMSS_package")
library(QMSS)
library(plm)

```

###1. Research Question

The following Directed Acyclic Graph (DAG) illustrates the relationships between variables assumed by the author, where "Out" is the Outcome of interest (being expenditures, revenues, and deficit of a municipality), "DD" Direct Democracy as a government system that affects government's spending: "FE" represents Fixed Effects within municipality and "u" an error term. Figure 1 reflects the same relationships modeled in Equation 1.


> **Equation 1.**   $Outcome_{myt} = \alpha_{m} + \beta DirDem_{mt} + f(Population_{mt} -100) + u_{myt}$


#### Figure 1. DAG for the research design of Direct democracy and government size: evidence from Spain

```{r}

#1. DAG of research design
#1.1 Generate coords vectors
coords1 <- list(
  x = c(DD = 0, Out = 0, Pop = 1, FE = .5, u = -0.5),
  y = c(DD = 0, Out = 1, Pop = 0, FE = 2, u = 2))

#1.2 Generate DAG
Figure1 <- dagify(Out ~ DD + Pop + FE + u,
               DD ~ Pop,
              coords = coords1)
ggdag(Figure1, layout = "circle") +
  theme_dag_blank()

```


```{r}
#1. Prepare data

library(haven)
sanz_psrm <- read_dta("C:/Users/199278/Downloads/sanz_psrm.dta")
View(sanz_psrm)

#1.1 Rename data

dat <- sanz_psrm

#1.2 Prepare data
dat <- dat %>% drop_na(expenditures_c) #Drop observations with missing data

da2 <- dat %>% drop_na(log_expenditures_c) #Drop observations with missing data

dat <- dat %>% filter(year > 1987 & year <= 2011) #Drop observations out-of-sample years. 

#1.2.2 Generating centered population, running variable
dat$running = dat$population-99.5

```

### 2. Empirical Strategy 

The author argues that a discontinuity in the density of the population sizes within the municipalities is observed at the threshold to be a candidate or not for being able to implement Direct Democracy as a government system.  This discontinuity is illustrated in Figure 1 (Figure 2 in the original paper).

```{r}
#2.1 Plot histogram

Figure2 <- ggplot(data=dat, aes(dat$population )) + 
  geom_histogram(binwidth = 5) + theme_classic() + 
  labs(title = "Figure 2. Histogram of population sizes",
       subtitle = "An observation is a municipality-year. Bins are 5-inhabitant wide", x = "Population", y = "Observations", caption = "Adapted from Direct democracy and government size: evidence from Spain, by C. Sanz, 2017.") + geom_vline(xintercept = 100, linetype="dotted", 
                color = "white")

Figure2

```
Another visual inspection to check wether it makes sense to use a Regression Discontinuity Design (RDD) as the one used by the author is provided in Figure 3. 

```{r}
#2.2 Plot for visual check for RD
Figure3 <- dat %>% 
  ggplot(aes(x = running, y = as.factor(dd), col = as.factor(dd))) + geom_point() +
  labs(title = "Figure 3.Running variable and government system",
       x = "Running variable (Population size, centerd around 100 inhabitants)",
       y = "Government system") +
  scale_colour_discrete(name="Government\nSystem",
                          breaks=c("0", "1"), 
                        labels=c("Representative", "Direct")) +
  theme_classic()

Figure3

```

### 3. Main Results


Sanz examines the effect of direct democracy on public finances in Spanish municipalities by studying three different variables: Expenditures (log), Revenues (log) and Deficit. He uses seven different panel specifications for *Equation 1* and I replicate the estimates of the first four columns of Table 2 (original paper) for the three panels. In all regressions, fixed time and unit-specific effects are included. Each column of Table 1(replication) reports the estimated coefficients, standard errors (clustered at municipality level), number of observations, type of specification and bandwidth used for the linear model for panel data.  I estimate the optimal bandwidth for each panel as proposed by Imbens and Kalyanaraman (2012) using a rectangular kernel.

The main findings show that direct democracy has a statistically significant effect on finances by decreasing the size of government measured as the total log expenditures and log revenues. The effect of direct democracy has no significant effect on the overall deficits of municipalities. As it is shown in Table 1, it reduces public expenditure by almost 8 percent and public revenues by 5 percent (column 1). The reported estimations of the coefficients for all three panels and all 4 specifications are presented in Table 1. 

```{r}
#3.1 Generate bandwidths with IK method

#Bandwidth Deficit
bw1 <- rdbwselect_2014(dat$deficit_c, dat$running, c = 0, 
                       kernel = "uni", bwselect = "IK")

#Bandwidth Reveneues
bw2 <- rdbwselect_2014(dat$log_revenues_c, dat$running, c = 0, 
                       kernel = "uni", bwselect = "IK")

#Bandwidth Expenditures
bw3 <- rdbwselect_2014(dat$log_expenditures_c, dat$running, c = 0, 
                       kernel = "uni", bwselect = "IK")

#3.2 Rename bandwidths and extract values

bw_ex <- bw3$bws[1]
bw_rev <- bw2$bws[1]
bw_def <- bw1$bws[1]
```

```{r}
#3.3 Run panel data linear models 

#Panel A. Log Expenditures

dat$ddpop <- dat$running*dat$dd #generate interaction variable

#Column 1: Linear specification, BW=27.94

plm_ex <- plm(log_expenditures_c ~ running + dd + ddpop +as.factor(year), 
               data = subset(dat, running > 0 - bw_ex & running < 0 + bw_ex), 
               index = c("id"), model = "within")
plm_ex.ct <- coeftest(plm_ex, vcov. = vcovHC(plm_ex, type="sss", cluster="group")) #Clustered SE


#Column 2: Linear specification, BW=27.94*1.5

bw_ex15 <- (bw_ex*1.5)

plm_ex15 <- plm(log_expenditures_c ~ running + dd + ddpop +as.factor(year), 
                 data = subset(dat, running > 0 - bw_ex15 & running < 0 + bw_ex15), 
                 index = c("id"), model = "within")
plm_ex15.ct <- coeftest(plm_ex15, vcov. = vcovHC(plm_ex15, type="sss", cluster="group"))

#Column 3: Linear specification, BW=27.94*.5
bw_ex5 <- (bw_ex*.5)


plm_ex5 <- plm(log_expenditures_c ~ running + dd + ddpop + as.factor(year), 
                data = subset(dat, running > 0 - bw_ex5 & running < 0 + bw_ex5), 
                index = c("id"), model = "within")
plm_ex5.ct <- coeftest(plm_ex5, vcov. = vcovHC(plm_ex5, type="sss", cluster="group"))

#Column 4: Polynomial specification, full

plm_ex3 <- plm(log_expenditures_c ~ dd + poly(running, 3, raw = T) +
                 poly(running, 3, raw = T)*dd, 
                data = dat, 
                index = c("id", "year"), model = "within",
               effect = "twoways")
plm_ex3.ct <- coeftest(plm_ex3, vcov. = vcovHC(plm_ex3, type="sss", cluster="group"))

#***********************************************************************

#Panel B. Log Revenues
#Column 1: Linear specification, BW=25.42

plm_rev <- plm(log_revenues_c ~ running + dd + ddpop +as.factor(year), 
                data = subset(dat, running > 0 - bw_rev & running < 0 + bw_rev), 
                index = c("id"), model = "within")
plm_rev.ct <- coeftest(plm_rev, vcov. = vcovHC(plm_rev, type="sss", cluster="group"))

#Column 2: Linear specification, BW=25.42*1.5

bw_rev15 <- (bw_rev*1.5)


plm_rev15 <- plm(log_revenues_c ~ running + dd + ddpop +as.factor(year), 
                  data = subset(dat, running > 0 - bw_rev15 & running < 0 + bw_rev15), 
                  index = c("id"), model = "within")
plm_rev15.ct <- coeftest(plm_rev15, vcov. = vcovHC(plm_rev15, type="sss", cluster="group"))

#Column 3: Linear specification, BW=25.42*.5
bw_rev5 <- (bw_rev*.5)


plm_rev5 <- plm(log_revenues_c ~ running + dd + ddpop +as.factor(year), 
                 data = subset(dat, running > 0 - bw_rev5 & running < 0 + bw_rev5), 
                 index = c("id"), model = "within")
plm_rev5.ct <- coeftest(plm_rev5, vcov. = vcovHC(plm_rev5, type="sss", cluster="group"))

#Column 4: Polynomial specification, full

plm_rev3 <- plm(log_revenues_c ~ dd + poly(running, 3, raw = T) +
                 poly(running, 3, raw = T)*dd, 
                data = dat, 
                index = c("id", "year"), model = "within",
               effect = "twoways")
plm_rev3.ct <- coeftest(plm_rev3, vcov. = vcovHC(plm_rev3, type="sss", cluster="group"))

#*************************************************************************

#Panel C. Deficit (euros)
#Column 1: Linear specification, BW=98.60

plm_defm <- plm(deficit_c ~ running + dd + ddpop +as.factor(year), 
               data = subset(dat, running > 0 - bw_def & running < 0 + bw_def), 
               index = c("id"), model = "within")
plm_defm.ct <- coeftest(plm_defm, vcov. = vcovHC(plm_defm, type="sss", cluster="group"))

#Column 2: Linear specification, BW=98.60*1.5

bw_def15 <- (bw_def*1.5)

plm_defm15 <- plm(deficit_c ~ running + dd + ddpop +as.factor(year), 
                data = subset(dat, running > 0 - bw_def15 & running < 0 + bw_def15), 
                index = c("id"), model = "within")
plm_defm15.ct <- coeftest(plm_defm15, vcov. = vcovHC(plm_defm15, type="sss", cluster="group"))

#Column 3: Linear specification, BW=98.60*.5

bw_def5 <- (bw_def*.5)

plm_defm5 <- plm(deficit_c ~ running + dd + ddpop +as.factor(year), 
                  data = subset(dat, running > 0 - bw_def5 & running < 0 + bw_def5), 
                  index = c("id"), model = "within")
plm_defm5.ct <- coeftest(plm_defm5, vcov. = vcovHC(plm_defm5, type="sss", cluster="group"))

#Column 4: Polynomial specification, full
plm_def3 <- plm(deficit_c ~ dd + poly(running, 3, raw = T) +
                 poly(running, 3, raw = T)*dd, 
                data = dat, 
                index = c("id", "year"), model = "within",
               effect = "twoways")
plm_def3.ct <- coeftest(plm_def3, vcov. = vcovHC(plm_def3, type="sss", cluster="group"))

```

```{r}
#3.4 Preparate data for table 1

#3.4.1Extract Coefficients

#Panel A. Expenditures
plmA <- round(plm_ex$coefficients[2], digits = 4)
plmA15 <- round(plm_ex15$coefficients[2], digits = 4)
plmA5 <- round(plm_ex5$coefficients[2], digits = 4)
plmA3 <- round(plm_ex3$coefficients[1], digits = 4)

#Panel B. Revenues
plmB <- round(plm_rev$coefficients[2], digits = 4)
plmB15 <-round(plm_rev15$coefficients[2], digits = 4)
plmB5 <-round(plm_rev5$coefficients[2], digits = 4)
plmB3 <-round(plm_rev3$coefficients[1], digits = 4)

#Panel C. Deficit
plmC <-round(plm_defm$coefficients[2], digits = 3)
plmC15 <-round(plm_defm15$coefficients[2], digits = 3)
plmC5 <-round(plm_defm5$coefficients[2], digits = 3)
plmC3 <-round(plm_def3$coefficients[1], digits = 3)

#3.4.2 Extract clustered standard errors

#Panel A. Expenditures
plmASE <-round(plm_ex.ct[2,2], digits = 4)
plmA15SE <-round(plm_ex15.ct[2,2], digits = 4)
plmA5SE <-round(plm_ex5.ct[2,2], digits = 4)
plmA3SE <-round(plm_ex3.ct[1,2], digits = 4)


#Panel B. Revenues
plmBSE <-round(plm_rev.ct[2,2], digits = 4)
plmB15SE <-round(plm_rev15.ct[2,2], digits = 4)
plmB5SE <-round(plm_rev5.ct[2,2], digits = 4)
plmB3SE <-round(plm_rev3.ct[1,2], digits = 4)

#Panel C. Deficit
plmCSE <-round(plm_defm.ct[2,2], digits = 3)
plmC15SE <-round(plm_defm15.ct[2,2], digits = 3)
plmC5SE <-round(plm_defm5.ct[2,2], digits = 3)
plmC3SE <-round(plm_def3.ct[1,2], digits = 3)


#3.4.3 Extract number of observations

plm_An <- nobs(plm_ex)
plm_A15n <- nobs(plm_ex15)
plm_A5n <- nobs(plm_ex5)
plm_A3n <- nobs(plm_ex3)

plm_Bn <- nobs(plm_rev)
plm_B15n <- nobs(plm_rev15)
plm_B5n <- nobs(plm_rev5)
plm_B3n <- nobs(plm_rev3)

plm_Cn <- nobs(plm_defm)
plm_C15n <- nobs(plm_defm15)
plm_C5n <- nobs(plm_defm5)
plm_C3n <- nobs(plm_def3)
```

####Table 1. Effect of Direct Democracy on Public Finances
```{r}

#3.5 Generate table 1
table1 <- matrix(c(plmA,plmA15,plmA5,plmA3,
                   plmASE,plmA15SE,plmA5SE,plmA3SE ,
                   plm_An,plm_A15n,plm_A5n,plm_A3n,
                   "Linear", "Linear", "Linear", "Order 3",
                   "Optimal a", "Optimal a x 1.5", "Optimal a x 0.5", "Full",
                  plmB,plmB15,plmB5,plmB3,
                  plmBSE,plmB15SE,plmB5SE,plmB3SE ,
                  plm_Bn,plm_B15n,plm_B5n,plm_B3n,
                  "Linear", "Linear", "Linear", "Order 3",
                  "Optimal b", "Optimal b x 1.5", "Optimal b x 0.5", "Full",
                  plmC,plmC15,plmC5,plmC3,
                  plmCSE,plmC15SE,plmC5SE,plmC3SE ,
                  plm_Cn,plm_C15n,plm_C5n,plm_C3n,
                  "Linear", "Linear", "Linear", "Order 3",
                  "Optimal c", "Optimal c x 1.5", "Optimal c x 0.5", "Full"),ncol=4,byrow=TRUE)

colnames(table1) <- c("( 1 )","( 2 )","( 3 )","( 4 )")
rownames(table1) <- c("Direct Democracy","SE","Observations", "Specification", "Bandwidth",
                     "Direct Democracy","SE","Observations", "Specification", "Bandwidth",
                     "Direct Democracy","SE","Observations", "Specification", "Bandwidth")
table1 <- as.table(table1)


#3.6 Format table 1

kable(table1) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", position = "float_center")) %>%
  footnote(general = "Colums 1-4 report estimated coefficients for Equation (1) for all panels with a uniform kernel including time and unit fixed effects. This table was adapted from Direct democracy and government size: evidence from Spain, by C. Sanz, 2017.",
           number = c("Optimal BW a = 27.94, Optimal BW b = 25.42, Optimal BW c = 98.60", 
                      "Standar Errors, clustered at municipality level.", 
                      "Discrepancies in Standar Errors and number of observations might be due to the decimal points difference between the bandwidths used in the paper and the ones I used to make the estimations."))%>%
  pack_rows("Panel A. Log Expenditures", 1, 5, label_row_css = "background-color: #666; color: #fff;")  %>%
  pack_rows("Panel B. Log Revenues", 6, 10, label_row_css = "background-color: #666; color: #fff;")  %>%
  pack_rows("Panel C. Deficit (euros)", 11, 15, label_row_css = "background-color: #666; color: #fff;") 

```
***
In Figure 3 of his paper, the autor estimates  $Outcome_{myt} = \alpha_{m} + \gamma_{y}+ \sum_{100-OBW}^{100+OBW} \delta_{j}Population_{j, mt} + u_{myt}$ where $Population_{j, mt}$ is a dummy variable that indicates if municipality $m$ has population size $j$ at term $t$. In the y axis, estimated coefficients $\delta_{j}$. 

I use several data-driven RD plots that are evenly-spaced with binned sample means tracing out the underlying regression discontinuity function (Equation 1) using a rectangular kernel. The variables are not standarized and in the following Figure I plot the running variable against the outcome of interest, which explains the difference with the original Figure 3  from Sanz'S paper. 

#### Figure 4. Effect of direct democracy on expenditures, revenues, and deficit. 
```{r}

#3.7 Plot the regressions

rdplot(dat$log_expenditures_c, dat$running, c = 0, nbins = 8, binselect = "esmv",
       kernel = "uni", p = 1, x.label = "Running Variable", 
       y.label = "Log Expenditures", title = "Effect of direct democracy on expenditures")

rdplot(dat$log_revenues_c, dat$running, c = 0, nbins = 8, binselect = "esmv",
       kernel = "uni", p = 1, x.label = "Running Variable", 
       y.label = "Log Revenues", title = "Effect of direct democracy on revenues")

rdplot(dat$deficit_c, dat$running, c = 0, nbins = 8, binselect = "esmv",
       kernel = "uni", p = 1, x.label = "Running Variable", 
       y.label = "Deficit", title = "Effect of direct democracy on Deficit")
```

###4. Discussion

####4.1 Modeling Strategy

*Assumptions*

The main identification assumption of the research paper is that there are no other variables that confound the effect of direct democracy on the outcome of interest -holding municipality and year fixed effects-.  

*Strenghts*

The strenghts identified in the empirical strategy are the several robustness checks that are run by the author and included in the appendix that support the research high internal validity. Also, the effect of direct democracy remains statistically significant throughout all the seven different specifications presented in the paper, which suggests that the design is not sensible to the form of the function selected to estimate such effect. The latter confirms the high internal validity of the study. 

*Limitations*

The research design relies strongly on the identification assumption and, even though the author carries out several robustness checks, the empirical strategy is population threshold dependent. This means that there can also be confounding and sorting. I argue that several factors are not considered in the research and that these might confound or generate bias in the estimation of the effect of direct democracy on the outcomes of interest. One of the main trends that have been present during the last three decades in Spain is the increasing depopulation of rural areas (i.e. municipalities under 100 inhabitants). Moreover, it could be also true that the demographic composition of the localities that are eligible for this type of government directly affects the outcome of interest given that on average, over 60% of their population is over 65 and without any inhabitants younger than 20. In the online appendix, the dicussion related to demographic variables that might influence the estimations does not consider this. 

However, considering that the sizes of the municipalities are approximately randomized at the exogenously assigned population thresholds, the use of a regression discontinuity design seems appropriate for the research and might ease the concerns regarding the other factors I proposed at the beginning of this section.  The latter would back the findings of this research paper for they are consistent with other research studies in Switzerland, the US, and New England (See  Besley et al. & Feld et al. 2003). Nonetheless, it is worth mentioning that the extrapolation of RD estimates away from the threshold sacrifices the design's advantage in internal validity (see de la Cuesta et. al 2016), which is why I present an alternative specification of the model in section 5.

####4.2 Robustness check 

As a robustness check, I conduct a regression discontinuity design method to account for observed covariates in the RD original design aiming to increase precision of treatment effect estimation. I argue that having direct democracy as a government system in the period t-1 also affects public finances. In table 2, I report the estimated coefficients and use the Optimal bandwidth that Sanz uses for figure 3 in his paper (Equation 2).

> **Equation 2.**   $Outcome_{myt} = \alpha_{m} + \beta_{1} DirDem_{mt}+ \beta_{2} DirDem_{m t-1} + f(Population_{mt} -100) + u_{myt}$

Direct democracy remains as statistically significant factor affecting both, expenditures and revenues in Spanish municipalities. However, it is also observed that the lagged treatment variable affects the outcome of interest in almost 3 and 4 percent in expenditures and revenues, respectively. This results are similar and consistent to the ones presented in the original paper and further contribute to the discussion that the effect of direct democracy on public finances increases over time.

####Table 2. Effect of Direct Democracy on Public Finances adjusted by having DD as a government system 
```{r, results = "asis"}

#4.1 Run panel data linear models 

#Panel A
plm_ex2 <- plm(log_expenditures_c ~ running*dd + l_dd, 
               data = subset(dat, running > 0 - bw_ex & running < 0 + bw_ex), 
               index = c("id", "year"), model = "within", effect = "twoways")
plm_ex2.ct <- coeftest(plm_ex2, vcov. = vcovHC(plm_ex2, type="sss", cluster="group")) #Cluster SE

#Panel B
plm_rev2 <- plm(log_revenues_c ~ running*dd + l_dd, 
               data = subset(dat, running > 0 - bw_ex & running < 0 + bw_ex), 
               index = c("id", "year"), model = "within", effect = "twoways")
plm_rev2.ct <- coeftest(plm_rev2, vcov. = vcovHC(plm_rev2, type="sss", cluster="group"))

#Panel C
plm_def2 <- plm(deficit_c ~ running*dd + l_dd, 
                data = subset(dat, running > 0 - bw_ex & running < 0 + bw_ex), 
                index = c("id", "year"), model = "within", effect = "twoways")
plm_def2.ct <- coeftest(plm_def2, vcov. = vcovHC(plm_def2, type="sss", cluster="group"))

#4.2 Extract data for table 2

#Coefficients DD
plmA2 <- round(plm_ex2$coefficients[2], digits = 4)
plmB2 <- round(plm_rev2$coefficients[2], digits = 4)
plmC2 <-round(plm_def2$coefficients[2], digits = 3)

#Coeff lagged DD
plmA2.l <- round(plm_ex2$coefficients[3], digits = 4)
plmB2.l <- round(plm_rev2$coefficients[3], digits = 4)
plmC2.l <-round(plm_def2$coefficients[3], digits = 3)


#SELagged DD
plmASE2.l <-round(plm_ex2.ct[3,2], digits = 4)
plmBSE2.l <-round(plm_rev2.ct[3,2], digits = 4)
plmCSE2.l <-round(plm_def2.ct[3,2], digits = 3)

#SE DD
plmASE2 <-round(plm_ex2.ct[2,2], digits = 4)
plmBSE2 <-round(plm_rev2.ct[2,2], digits = 4)
plmCSE2 <-round(plm_def2.ct[2,2], digits = 3)

#Number of observations
plm_A2n <- nobs(plm_ex2)
plm_B2n <- nobs(plm_rev2)
plm_C2n <- nobs(plm_def2)


#4.3 Create table 2
table2 <- matrix(c(plmA2,plmB2,plmC2,
                   plmASE2,plmBSE2,plmCSE2,
                   plmA2.l,plmB2.l,plmC2.l,
                   plmASE2.l,plmBSE2.l,plmCSE2.l,
                   plm_A2n,plm_B2n,plm_C2n,
                   "Linear", "Linear", "Linear",
                   "Optimal a", "Optimal a", "Optimal a"),ncol=3,byrow=TRUE)

colnames(table2) <- c("Log Expenditures","Log Revenues","Deficit (euros)")
rownames(table2) <- c("Direct Democracy","SE","Direct Democracy t-1","SE","Observations", "Specification", "Bandwidth")
table2 <- as.table(table2)


##4.4 Format table 2
kable(table2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", position = "center")) %>%
  footnote(general = "Colums 1-3 report estimated coefficients for Equation (2) for all panels with a uniform kernel including time and unit fixed effects. Optimal bandwidth calculated as suggested as in Imbens und Kalyanaraman 2012, using same bandwidth (Optimal a) for all three models, as Sanz's uses it in Figure 3 in the original paper.", 
                      "Standar Errors are presented as clustered at municipality level.") %>%
  add_header_above(c(" " = 1, "Panel A" = 1, "Panel B" = 1, "Panel C" = 1))

```


###5. Alternative Identification Strategy 

I propose a Difference in Differences model that takes into account the pre-treatment levels of the outcome of interest. I include a lagged variable of the treatment variable (direct democracy) as a regressor in the three panels as suggested in Wing et. al 2018. The coefficients substantially differ from those of the estimated with the RD design. The sign of the effect is positive in contrast to the estimations shown in the paper (yet, they remain statistically significant). 

The latter suggests that indeed, direct democracy might be a factor that has an effect on public finances but it might change as a function of the population and of time increasing the size of the govenrment.The explanatory power of the model is considerably lower than the one from the original design. This might be due to the observations used for the analysis. In the DiD estimation, even though there are observations dropped, over 95% of the observations are considered.

```{r, warning = FALSE, results = "asis"}

#5.1 Generate variables representing the change in outcome of interest
dat <- dat %>% 
  mutate(change_logex = log_expenditures_c - l_log_expenditures_c,
         change_logrev = log_revenues_c - l_log_revenues_c,
         change_def = deficit_c - l_deficit_c)

#5.2 Generate model diff in dif

did_A1 <- lm(log_expenditures_c ~ dd*l_dd, data = dat) #Panel A

did_B1 <- lm(log_revenues_c ~ dd*l_dd, data = dat) #Panel B

did_C1 <- lm(deficit_c ~ dd*l_dd, data = dat) #Panel C

stargazer(did_A1, did_B1, did_C1,type = 'html',digits = 3,
          title            = "Difference in Difference estimates for three panels",
          covariate.labels = c("Direct democracy", "Direct democracy t-1", "Interaction", "Constant"),
          dep.var.caption  = "Table 3. Difference in difference estimations of the effect of direct democracy on public finances",
          dep.var.labels   = c("Log Expenditures", "Log Revenues", "Deficit"))
```

Another reason that might explain the discrepancies between results might be because by including the  lagged treatment variable in the DID model we include Time-Varying Treatment Effects that the original plm model does not capture and this might also imply that the treatment effect could be heterogeneous among units. It is also worth noticing that we observe a statistical significative effect for the interaction of direct democracy in *t* and *t-1*. This might represent a hint of interest for further research on how direct democracy affects fiscal discipline over time.  Furthermore, it would also be interesting to try to collect data on the coalitions in the govenrment and include them in the analysis to examine its relation to fiscal discipline.

***

###References

Abraham, S., & Sun, L. (2018). Estimating Dynamic Treatment Effects in Event Studies With Heterogeneous Treatment Effects. SSRN Electronic Journal. doi: 10.2139/ssrn.3158747

Besley, T. and A. Case (2003). Political institutions and policy choices: Evidence from the United States. Journal of Economic Literature 41 (1), 7-73

De la Cuesta, B., & Imai, K. (2016). Misunderstandings About the Regression Discontinuity Design in the Study of Close Elections. Annual Review Of Political Science, 19(1), 375-396. doi: 10.1146/annurev-polisci-032015-010115

Feld, L. P. and J. G. Matsusaka (2003). Budget referendums and government spending: evidence from the Swiss cantons. Journal of Public Economics 87 (12), 2703-2724

Fr?lich, M., & Huber, M. (2018). Including Covariates in the Regression Discontinuity Design. Journal Of Business & Economic Statistics, 37(4), 736-748. doi: 10.1080/07350015.2017.1421544

Hlavac, Marek (2018). stargazer: Well-Formatted Regression and Summary Statistics Tables. R package version 5.2.1.

Sanz, C. (2019). Direct democracy and government size: evidence from Spain. Political Science Research And Methods, 1-16. doi: 10.1017/psrm.2018.65

Wing, C., Simon, K., & Bello-Gomez, R. (2018). Designing Difference in Difference Studies: Best Practices for Public Health Policy Research. Annual Review Of Public Health, 39(1), 453-469. doi: 10.1146/annurev-publhealth-040617-013507

